# 2 Assignment 2

<br>

### 2.1. Data Wrangling
After completing the Exercise set 2, I did the first part of the Assignment 2; Data wrangling. The first three steps were quite easy after the Exercise set, but setting the working directory by using write_csv() function was not so straightforward. I'm using a company computer and therefore I don't have a writing access to the folders by default. After a quick [google search](https://stackoverflow.com/questions/17779004/cannot-export-data-to-a-file-in-r-write-csv) I managed to overcome this with writing an absolute path to a file (see example below, not just the folder location). 

<br>

```
# Setting the working directory to the IODS Project folder using write_csv() function. 
write_csv(learning2014, file="C:\\...\\IODS\\IODS-project\\data\\learning2014.csv")
```

### 2.2. Analysis; Reading and exploring the data

Reading the data created in Data Wrangling part from a local folder:

```{r, warning=FALSE, message=FALSE}

# access all the libraries needed in the assignment 2:
library(dplyr)
library(tidyverse)
library(GGally)
library(ggplot2)


# Reading a .csv file created in the Data wrangling:
students2014 <- read_csv(file="C:\\Users\\kuud\\Work Folders\\IODS\\IODS-project\\data\\learning2014.csv")
```

<br>

Exploring the structure and dimensions of the data with **str()** and **dim()**. 
There are 166 observations and 7 variables in the data. 

Variables are:

1. gender; M (Male) and F (Female)\
2. Age; in years
3. Attitude; Global attitude toward statistic
4. deep; mean of the questions related to deep learning
5. stra; mean of the questions related to strategic learning
6. surf; mean of the questions related to surface learning
7. Points; Examination points

```{r, warning=FALSE, message=FALSE}
#structure of the data
str(students2014)

#dimensions of the data
dim(students2014)
```

<br> 

### 2.3. Analysis; Graphical overview of the data

Because there is only seven variables saved as columns in the data frame, I'll visualize all of the relationships in a single plots below using [**ggplot2**](http://ggplot2.org/) and **GGally** libraries. I found the [cheatsheet](https://rstudio.github.io/cheatsheets/html/data-visualization.html) for ggplot2 to be useful.


```{r, warning=FALSE, message=FALSE}
# draw a scatter plot matrix of the variables in students2014.
# [-1] excludes the first column (gender)
pairs(students2014[-1])

# create a plot matrix with ggpairs()
plot1 <- ggpairs(students2014, mapping = aes(), lower = list(combo = wrap("facethist", bins = 20))) + ggtitle ("The relationships of the variables")

# draw the plot
plot1

```

<br> 

As seen above, the **Attitude** seems to correlate the most with the Examination points. Let's examine the correlation a bit further:

```{r, warning=FALSE, message=FALSE}

# initialize plot with data, aesthetic mapping, define the visualization type (points), add regression line and a title
plot2 <- ggplot(students2014, aes(x = Attitude, y = Points)) + geom_point() + geom_smooth(method = "lm") + ggtitle ("Student's attitude versus Exam Points")

# draw the plot
plot2


```

<br> 

The other two variables that correlate the most with the Examination points are **stra** and **surf**. Let's examine them too:

```{r, warning=FALSE, message=FALSE}

# initialize plot with data, aesthetic mapping, define the visualization type (points), add regression line and a title
plot3 <- ggplot(students2014, aes(x = stra, y = Points)) + geom_point() + geom_smooth(method = "lm") + ggtitle ("Student's strategic learning versus Exam Points")
plot4 <- ggplot(students2014, aes(x = surf, y = Points)) + geom_point() + geom_smooth(method = "lm") + ggtitle ("Student's surface learning versus Exam Points")

# draw the plot
plot3
plot4


```

<br> 

### 2.4. Analysis; Regression model

The regression model below has three *explanatory variables*: **Attitude**, **stra** and **surf**. I chose the variables that had the highest absolute correlation with Examination **Points** (=*outcome variable*). 

```{r,warning=FALSE, message=FALSE}

# create a multiple regression model with explanatory variables: Attitude, stra and surf and outcome variable: Points
mreg_model <- lm(Points ~ Attitude + stra + surf, data = students2014)

# print out a summary of the model
summary(mreg_model)

```

Looking at the results: It seeems that Attitude explains the Exam Points the most (statistically significant p-value), which is in line with the plots seen above (the regression line has the strongest angle compared to the ones in plots visualizing the correlation between points and stra/surf). In other words, in the Coefficients section, **stra** and **surf** are not considered to be significant by the model because their p-value is higher than 0.05. Keeping them does not provide any additional value to the model, so regression model using only Attitude would be best.

<br>

The R-Square of the model means the proportion of variance explained by the model, which is 0.21 = the model explains 21 % of the Examination Points. If we use only Attitude in the model (below), the R-Square is 0.19 (not much changed when **stra** and **surf** removed from the model), which means Attitude towards statistics explains 19 % of the Examination points. Further we can see that for every one **Point** increase, the value of **Attitude** increases 0.35.

```{r,warning=FALSE, message=FALSE}

# create a multiple regression model with explanatory variable Attitude and outcome variable Points
reg_model <- lm(Points ~ Attitude, data = students2014)

# print out a summary of the model
summary(reg_model)

```

<br>

Finally, let's produce a few diagnostic plots, first the multiple regression model:

```{r,warning=FALSE, message=FALSE}

# draw diagnostic plots using the plot() function. Choose the plots 1, 2 and 5
par(mfrow = c(2,2))
plot(mreg_model, which=c(1,2,5))

```

<br>

... and then the regression model, using only **Attitude** as explanatory variable:

```{r,warning=FALSE, message=FALSE}

# draw diagnostic plots using the plot() function. Choose the plots 1, 2 and 5
par(mfrow = c(2,2))
plot(reg_model, which=c(1,2,5))

```
<br>

As seen above, the plots are very much similar when using only linear regression versus multiple regression. It seems that the attitude matters the most in this dataset. Interesting! :) 

<br>
<br>